---
title: "Stat380Assignment2"
author: "James DiPasquale, Tyler Gully, Ahmad"
format: html
---

## Introduction/Background

##### Insurance pricing depends heavily on understanding which factors actually drive cost. Nittany Valley Insurance Company provided us with policy data and asked us to build a predictive model to predict total insurance charges using linear regression techniques. The dataset includes demographic and behavioral variables such as age, sex, BMI, smoking status, number of dependents, region, claim month, and prior claim activity, along with the total charges billed to each policy.

##### Our objective is to use this information to build and evaluate predictive models for insurance charges. To do this, we begin with exploratory data analysis to better understand patterns, distributions, and relationships within the data. From there, we develop multiple candidate linear regression models, interpret the estimated coefficients, and assess model performance using appropriate validation methods.

##### The ultimate goal of this analysis is to recommend the best possible model that is both statistically sound and practically useful for the company’s pricing and risk management decisions.

## Data Exploration

```{r}
#| echo: false

library(ggplot2)

insurance_data <- read.csv("insurance_data_S26.csv", header = TRUE)

cols_num <- which(sapply(insurance_data, is.numeric))

xcont <- list("age", "bmi")

cols_num <- which(sapply(insurance_data, function (x) !(is.numeric(x))))

xcat <- list("smoker", "children", "priorclaims")

# Excluded Variables: X, caseid, month, sex, region

print(xcont)
print(xcat)

for (col_name in xcont) {
  plot <- ggplot(insurance_data, aes(insurance_data[[col_name]], charges)) + 
  geom_point() +
    geom_smooth(data = insurance_data, method = "lm", color = "red") +
    labs(
    title = paste(col_name, " vs Charges"),
    x = paste(col_name),
    y = "Charges"
  )
  print(plot)
}

for (col_name in xcat) {
  plot <- ggplot(insurance_data, aes(as.factor(insurance_data[[col_name]]), charges)) + 
  geom_boxplot() +
    labs(
    title = paste(col_name, " vs Charges"),
    x = paste(col_name),
    y = "Charges"
  )
  print(plot)
}

```


##### During Exploratory Data Analysis, we used a for loop to graph each predictor variable against charges, the response variable we are trying to predict. After reviewing the initial plots, we determined that categorical variables would be better visualized using boxplots, while continuous variables were more appropriately displayed as scatterplots. We therefore separated the categorical and continuous variables, plotting the categorical variables as boxplots and the continuous variables as scatterplots with a fitted linear regression line to better assess potential relationships.

## Modeling Methods/Strategies

```{r}

# col_list = columns identified as valid
col_list <- list("age", "bmi", "smoker", "children", "priorclaims")

# Initial Model selection using backwards AIC/BIC using the col_list

# Full model with all predictors
full_model <- lm(charges ~ age + bmi + smoker + children + priorclaims,

                 data = insurance_data)

summary(full_model)

# Stepwise selection using AIC
step_model <- step(full_model, direction = "both")

summary(step_model)

# Compare AIC values
AIC(full_model)
AIC(step_model)
```

## Results

### A. Initial Results

```{r}
# Show summaries for both models
summary(full_model)
summary(step_model)

# Compare model fit statistics
AIC(full_model)
AIC(step_model)
BIC(full_model)
BIC(step_model)
```

##### The initial regression results showed that the model worked pretty well overall and explained a large amount of the variation in insurance charges. With an $R^2$ of 0.7222, the model explained about 72.22% of the variability in charges, which means the predictors did a good job capturing what affects cost. Smoking had by far the biggest impact on charges, while age, BMI, number of children, and prior claims also all had positive effects. Since the stepwise model ended up with the exact same AIC and BIC values as the full model, there was no reason to simplify it further, so we kept the full model.


### B. Interpretation & Model Evaluation

##### Looking at the model, all of the predictors had a positive effect on insurance charges. That means as age, BMI, number of children, and prior claims go up, the predicted charges also go up, assuming the other variables stay the same. Smoking had by far the biggest effect in the model, which shows that it is the strongest factor related to higher insurance costs. The other variables still mattered, but not nearly as much as smoking.

##### Overall, the model performed pretty well and explained a good amount of the variation in charges. With an $R^2$ of 0.7222 and an adjusted $R^2$ of 0.7205, the model explained about 72% of the variability in insurance charges, which suggests it did a solid job. The overall model was also highly statistically significant, so the predictors were useful when taken together. Since the stepwise model had the exact same AIC and BIC values as the full model, there was no real benefit to simplifying it, so we kept the full model.

### C. Refinements

```{r}

# Refined Model

# ...and Final Results

```

##### [Refinements Discussion]

### Final recommendations

##### A paragraph to summarize your recommended model and key findings.

